Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-03-27 transmission.tex
ch 7 proofing

In this chapter, I use the estimate of the duration of RT-PCR positivity from combining information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate the feasibility of estimating incidence and transmission of SARS-CoV-2 from only the CIS data (described in \cref{E-intro:sec:cis}).
I will show that the CIS data are sufficient, demonstrating the value of surveys of this type for pandemic surveillance.

@@

First, in \cref{backcalc}, I use a backcalculation approach, based on \cref{E-inc-prev:eq:EPt}.
This approach enables estimation of incidence within (sub)populations of England.
There are several attractive features about this approach in the context here.
The backcalculation approach makes few assumptions about how transmission occurs; the assumptions have the potential to introduce bias.
It can also easily incorporate arbitrary covariates to adjust for non-response bias.
The scale of the CIS and desire to correct for non-response bias presents challenges for this methodology, especially within the limited computation of the SRS environment.

@@

An important consideration in this section is the non-response bias.
I adopt the poststratification method introduced in \cref{E-biology-data:sec:MRP} to correct for this bias.
Therefore, incidence must be estimated in each stratum before being poststratified to the target population and subpopulations of interest.
The strata are defined by at least age group, region, ethnicity, and sex due to differential non-response in the survey (see \cref{E-intro:sec:cis}).

@@

In \cref{E-inc-prev:sec:observation-process}, I showed that a deterministic approximation to the relationship between true prevalence and incidence is a good approximation for estimating incidence.
This relationship between incidence proportion and prevalence means that once one is inferred, the other is calculated via a deterministic function.
This gives two possible approaches for estimating incidence from prevalence data.

@@

The first approach, which I call the \emph{incidence-inferred} approach, is a single statistical model where incidence is inferred, with prevalence as a derived quantity.
The basis of the statistical model, in a Bayesian framework, would be as follows.
\begin{align}
    \vec{Z} &\dist F_Z \\
    P_{i,t}  &= \sum_{k=0}^{d_\text{max}-1} Z_{i,t-k} S(k+1) \\
    \rho &\dist F_\rho \\
    Y_{i,t} \mid \vec{P}, \rho &\dist \BB(n_{i,t}, P_{i,t}, \rho)
\end{align}
where $\vec{Z}$ is a vector such that $Z_{i,t}$ is the incidence proportion in stratum $i$ on day $t$, with a joint prior $F_Z$; $P_{i,t}$ is the prevalence in stratum $i$ on day $t$; $\rho$ is the overdispersion parameter of the beta-binomial distribution with prior $F_\rho$; $Y_{i,t}$ is the number of positive tests in stratum $i$ on day $t$; and $n_{i,t}$ is the number of tests in stratum $i$ on day $t$.
Fitting this model to a prevalence survey such as the CIS (\eg using MCMC) gives a posterior distribution of $\vec{Z}$, the target of inference.

The relationship between $P$ and $Z$ is from \cref{E-inc-prev:eq:EPt}, although dividing both sides by the stratum's population (assumed constant) to have the values in terms of incidence proportion and prevalence rather than incidence and number of prevalent individuals.

@@

The second approach, which I call the \emph{prevalence-inferred} approach takes a two-step approach.
First, it uses the statistical model from \cref{E-biology-data:sec:MRP} to estimate prevalence.
Then, it transforms the posterior distribution of prevalence to a posterior distribution of incidence using the forward substitution as in \cref{inc-prev:eq:forward-substitute}.
At a high-level, the statistical model is:
\begin{align}
    \vec{P}  &\dist F_P \\
    \rho &\dist F_\rho \\
    Y_{i,t} \mid \vec{P}, \rho &\dist \BB(n_{i,t}, P_{i,t}, \rho)
\end{align}
where $F_P$ is analogous to $F_Z$ in providing a joint prior for the prevalence in each stratum and day combination, and all other notation is as in the incidence-inferred approach.
The first step is to fit this model, giving a posterior distribution for $\vec{P}$.

The second step is to transform the posterior distribution for $\vec{P}$ to a posterior distribution for $\vec{Z}$.
A posterior sample for $\vec{P}$ can be transformed to a posterior sample for $\vec{Z}$ based on \cref{inc-prev:eq:forward-substitute}, see \cref{backcalc:sec:methods} for details.

\subsection{Desirable properties}

The incidence-inferred approach is more parsimonious in being a single step procedure.
However, it poses several pragmatic challenges which means I favour the prevalence-inferred approach.

@@

The chosen approach should have the following properties.
\begin{enumerate}
    \item The prior should favour $\pi_{i,t}$ changing exponentially in $t$, \ie penalize changes from exponential growth.
        % Penalizing changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
        Such penalization has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately} and is the approach used for the CIS prevalence model (see \cref{E-biology-data:sec:MRP}).
    \item Limit estimates to epidemiologically and biologically plausible values.
        That is, incidence proportion and prevalence should be in the interval $[0, 1]$.
        % While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias.
    \item Borrow information across the different covariates.
        The borrowing should penalize changes in the distribution of incidence/prevalence across strata in the same geographic region.
        A constant distribution is predicted by the theoretical study of epidemics, for example the dynamics of stratified, mechanistic transmission models (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) over short periods of time with negligible behavioural and immunity changes.
    \item Be computationally efficient.
        The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see \cref{E-biology-data:sec:SRS}).
\end{enumerate}

A parsimonious model satisfying properties 1 and 2 is to specify $\logit(\pi_{i,t}) = g_\varphi(i, t)$, where $g$ is a function parameterised by $\varphi$ and the prior on $\varphi$ is specified to favour $d^2g/dt^2 = 0$, \ie favour linear growth in $\logit(\pi_{i,t})$.
The prior $F_Z$ or $F_P$ is now a prior on $\varphi$.
The $\logit$ link function both limits $\pi_{i,t}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{i,t}$~\autocite{ealesAppropriately}.

The logit transformation also allows property 4 to be conveniently satisfied using an additive model for $g_\varphi$.
An additive model on the logit scale is approximately multiplicative on the natural scale for small values of $\pi_{i,t}$ (which will be common).

@@

Properties 3 and 5 suggest the use of an approximate method for computing the posterior distribution.
In particular, using MCMC, sufficient covariates to adjust for the non-response bias, over a period of more than 6--8 weeks leads to a model that is computationally prohibitive ~\citePersonalComms{Koen Pouwels}.
Instead, previous MRP implementations for the CIS used INLA (see \cref{E-biology-data:sec:MRP}).

INLA-based inference can only be applied to LGMs (\emph{Latent Gaussian Models}); see \cref{E-transmission:sec:INLA} for further details on INLA and LGMs.
The LGM framework requires that a transformation of $P_{i,t}$ is linear in a vector of $g_{\varphi}(i, t)$s combining all $i$s and $t$s.

If $\pi_{i,t} = P_{i,t}$ then this is trivially satisfied.
However, if $\pi_{i,t} = Z_{i,t}$, then no such linear relationship exists.
In particular:
\begin{align}
    P_{i,t}
    &= \sum_{k=0}^{\dmax} \pi_{i,t-k} S(k+1) \\
    &= \sum_{k=0}^{\dmax} \expit(g_\varphi(i, t-k)) S(k+1).
\end{align}

@@

Now I explain how I extended the MRP model introduced in \cref{E-biology-data:sec:MRP} to implement the prevalence-inferred approach.
I start with the prevalence posterior from \cref{E-biology-data:sec:positivity-results}, denoting the times the estimates are for as $t = 1, \dots, T$.
I draw 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).

Then, I transform this posterior sample using \cref{transmission:alg:calculate-Z}.
The algorithm is based on \cref{inc-prev:eq:forward-substitute}, which is repeated here for convenience:
\begin{align}
Z_t
% &= \frac{P_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
&= P_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
\label{transmission:eq:forward-substitute}
\end{align}
Dividing each side of \cref{transmission:eq:forward-substitute} by the total number of individuals in each stratum transforms the number of prevalent individuals to prevalence and incidence to incidence proportion.
The algorithm uses a deterministic relationship between incidence and prevalence within a stratum, as \cref{E-inc-prev:sec:observation-process} justifies; \ie, the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.

@@

For each posterior sample of $\vec{Z}$, I poststratify it to the CIS target population and subpopulations of interest.
This is done for each posterior sample using the methodology explained in \cref{E-biology-data:sec:MRP}.
The poststratified posterior samples are now posterior samples from the posterior of interest.

@@

A slow increase in incidence is estimated to occur over the autumn 2020 period, peaking in October in most regions (see \cref{transmission:fig:backcalc-regions}).
An exception is the North East and East of England regions (see \cref{transmission:fig:backcalc-regions}(C) and (D) respectively); these regions show no consistent increase or decrease in incidence during October.
All regions then decrease during November; this coincides with the national lockdown which came into force in England on 5 Nov 2020~\autocite{ifgTimeline}.

During November, incidence starts rising, with regional heterogeneity in the timing of when this occurs (see \cref{transmission:fig:backcalc-regions}).
The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
There is evidence that when incidence starts increasing is correlated with the timing of the Alpha variant's spread (see \cref{transmission:fig:backcalc-alpha}).
A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
\begin{figure}
    \includegraphics{transmission/backcalc-regions}
    \caption[Incidence estimated using backcalculation by region]{%
        Incidence proportion for (A) all of England and (B) its component regions.
    }
    \label{transmission:fig:backcalc-regions}
\end{figure}
\begin{figure}
    \includegraphics{transmission/backcalc-alpha}
    \caption[Correlation of the emergence of Alpha and minimum incidence.]{%
        x-axis: time that S-gene target failure, a marker of the Alpha variant, is estimated to start increasing from baseline by region among CIS test results; from \textcite[table S3]{walkerTracking}.
        y-axis: posterior median and 95\% CrI of the date of minimum incidence in November or early December in each region.
        Linear regression line shown.
    }
    \label{transmission:fig:backcalc-alpha}
\end{figure}

@@

An edge effect, a rapid rise in incidence between days one and two, is visible (see \cref{transmission:fig:backcalc-start-effect}).
This is due to the incorrect assumption that incidence is constant in the past.
Changing this assumption (\eg to exponential growth) modifies the estimated value on the first day, but the rest of the series is insensitive to this change (not shown).
Therefore, the first day's estimate is unreliable, and has been excluded from all other results.
However, the rest of the series is robust to this assumption and hence reliable.
\begin{figure}
    \centering \includegraphics{transmission/backcalc-start-effect}
    \caption[Edge effects in backcalculation method]{%
        First seven days of incidence proportion estimate for the whole of England, highlighting the edge effect.
    }
    \label{transmission:fig:backcalc-start-effect}
\end{figure}

@@

The second set of assumptions are in specifying the prior for $\vec{P}$.
In particular, the prior places non-zero mass on prevalence values which imply negative incidence under the other model assumptions.
Rectifying this assumption is challenging without switching to the inference-inferred approach, discussed further in \cref{backcalc:sec:results}.

Alternative inference techniques could allow more powerful modelling approaches, if they allowed a broader set of models than LGMs.
In particular, the incidence-inferred approach could be adopted.
This could allow more interpretable priors to be used because the infection process is the process that corresponds to biological and behavioural processes.
Furthermore, it would prevent the issues with negative incidence.

@@

I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcite{vanleeuwenTime,vanleeuwenAugmenting}.
%; these estimates have been used for previous modelling of the SARS-CoV-2 pandemic~\autocite{vanleeuwenTime,birrellRealtime}.
% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
This method starts with pre-pandemic baseline estimates of $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
UKTUS asked a representative sample of the UK population about how much time they spent doing various activities (\eg shopping and spending time at home) each day.
\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
Google mobility data are based on the location of Android users over time, relative to a pre-pandemic baseline.
The mobility and school data are averaged over a week down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
$\kappa_{aa'}(t)$ is assumed to be piecewise constant, changing only at the start of each week.
For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.

@@

The time-varying component, $\beta(t)$, allows for behavioural changes that impact transmission but are not captured by the contact matrices (\eg mask wearing).
I impose a random walk prior on $\log\beta(t)$ on the weekly timescale, smoothing its value over time.
This prior encodes that behaviour changes smoothly over time.
\begin{align}
    \beta(t) &= \beta_{\tilde{w}(t)} \\
    \log\beta_{w+1} &= \log\beta_w + \sigma_\epsilon \epsilon_w \\
    \epsilon_w &\sim \N(0, 1)
\end{align}
where $\tilde{w}(t)$ is the week number of time $t$, with week 1 being the first week considered.
Two hyperparameters will be inferred: $\beta_1$, the initial value of the random walk; and $\sigma_\epsilon$, the standard deviation of the random walk (see \cref{SEIR:sec:priors}).
$\beta(t)$ is the only time-varying part of the transmission model that is not fixed a priori.
Therefore, in practice, it will also absorb other effects such as any error in $\matr{\kappa}$.

@@

The remaining details of the transmission model structure follow \textcite{birrellRealtime}.
The transmission model is an SEIR model with two latent and two infectious states.
To form whole of England estimates, each region is modelled independently with the results aggregated, using the poststratification methodology explained in \cref{E-backcalc}.

A diagram of the model for a single stratum is shown in figure \cref{SEIR:fig:full-model}.
The full system of differential equations for each region is below.
\begin{equation}
\begin{aligned}
    \vec{\lambda}(t) &= \beta(t) (\matr{\kappa}(t) \circ \matr{q} \circ \matr{N}) ((\vec{i_1}(t) + \vec{i_2}(t))) \\
    \frac{d\vec{s}(t)}{dt} &= -\vec{\lambda}(t) \circ \vec{s}(t) \\
    \frac{d\vec{e_1}(t)}{dt} &= \vec{\lambda}(t) \circ \vec{s}(t) - 2\sigma \vec{e_1}(t) \\
    \frac{d\vec{e_2}(t)}{dt} &= 2\sigma \vec{e_1}(t) - 2\sigma \vec{e_2}(t) \\
    \frac{d\vec{i_1}(t)}{dt} &= 2\sigma \vec{e_2}(t) - 2\gamma \vec{i_1}(t) \\
    \frac{d\vec{i_2}(t)}{dt} &= 2\gamma \vec{i_1}(t) - 2\gamma \vec{i_2}(t) \\
    \frac{d\vec{r}(t)}{dt} &= 2\gamma \vec{i_2}(t) \\
    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \omega_0 \vec{p_0}(t) \\
    \frac{d\vec{p_1}(t)}{dt} &= \omega_0 \vec{p_0}(t) - \omega_1 \vec{p_1}(t) \\
    \frac{d\vec{p_2}(t)}{dt} &= \omega_1 \vec{p_1}(t) - \omega_{2} \vec{p_2}(t) \\
    \frac{d\vec{p_3}(t)}{dt} &= \chi \omega_{2} \vec{p_2}(t) - \omega_{3} \vec{p_3}(t)
\end{aligned}
\label{SEIR:eq:fullODEs}
\end{equation}
where $\circ$ indicates element-wise multiplication and $\matr{N}$ is a matrix with $(\matr{N})_{aa'} = N_{a'} / N_{a}$.
See \cref{SEIR:eq:fullODEs}

@@

Using the results of \cref{E-SEIR:sec:structured-populations}, $\R$ can also be written as $R_{\beta} = \beta_1 D(\matr{\kappa}(t) \circ \matr{q} \circ \matr{N})$ (recall that $D(\matr{\beta})$ denotes the dominant eigenvalue of $\matr{\beta}$).
Hence setting $\beta_1 = R_\psi / D(\matr{\kappa}(t) \circ \matr{q} \circ \matr{N})$ gives $R_\psi = R_\beta$.
Using this expression, $\beta_1$ has been reparameterized in terms of $\psi$, $q_c$, and fixed quantities.

>>>>

In this chapter, I use the estimate of the distribution of the duration of RT-PCR positivity from combining information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate the feasibility of estimating incidence and transmission of SARS-CoV-2 from only the CIS data (described in \cref{E-intro:sec:cis}).
I will show that the CIS data are sufficient, demonstrating the value of surveys of this type for pandemic surveillance.

@@

I start with the backcalculation (\cref{backcalc}), based on \cref{E-inc-prev:eq:EPt}.
This approach enables estimation of incidence within (sub)populations of England.
There are several attractive features about this approach in the context here.
The backcalculation approach makes few assumptions about how transmission occurs; stronger assumptions increase the potential for introduce bias.
It can also easily incorporate arbitrary covariates to adjust for non-response bias.
The scale of the CIS and desire to correct for non-response bias presents challenges for this methodology, especially within the limited computation of the SRS environment.

@@

An important consideration in this section is non-response bias.
I correct for the non-response through the poststratification method introduced in \cref{E-biology-data:sec:MRP}.
Poststratification requires estimates of incidence in each stratum, before combining them to the target population and subpopulations of interest.
The strata are defined by at least age group, region, ethnicity, and sex due to differential non-response in the survey (see \cref{E-intro:sec:cis}).

@@

In \cref{E-inc-prev:sec:observation-process}, I showed that a deterministic approximation to the relationship between true prevalence and incidence is a good approximation for estimating incidence.
This relationship between incidence proportion and prevalence means that once one is inferred, the other is calculated via a deterministic function.
This gives two possible approaches for estimating incidence from prevalence data; in \cref{transmission:sec:backcalc-properties} I argue the second approach is more appropriate in this context.

@@

The first approach, which I call the \emph{incidence-inferred} approach, is a single statistical model where incidence is inferred, with prevalence as a derived quantity.
The basis of the statistical model, in a Bayesian framework, would be as follows.
\begin{align}
    \vec{Z} &\dist F_Z \\
    P_{i,t}  &= \sum_{k=0}^{d_\text{max}-1} Z_{i,t-k} S(k+1) \\
    \rho &\dist F_\rho \\
    Y_{i,t} \mid \vec{P}, \rho &\dist \BB(n_{i,t}, P_{i,t}, \rho)
\end{align}
where $\vec{Z}$ is a vector such that $Z_{i,t}$ is the incidence proportion in stratum $i$ on day $t$, with a joint prior $F_Z$; $P_{i,t}$ is the prevalence in stratum $i$ on day $t$; $\rho$ is the overdispersion parameter of the beta-binomial distribution (see \cref{E-distributions}) with prior $F_\rho$; $Y_{i,t}$ is the number of positive tests in stratum $i$ on day $t$; and $n_{i,t}$ is the number of tests in stratum $i$ on day $t$.
Fitting this model to a prevalence survey such as the CIS (\eg using MCMC) gives a posterior distribution of $\vec{Z}$, the target of inference.

The relationship between $\vec{P}$ and $\vec{Z}$ is from \cref{E-inc-prev:eq:EPt}, although dividing both sides by the stratum's population (assumed constant) to have the values in terms of incidence proportion and prevalence rather than incidence and number of prevalent individuals.

@@

The second approach, which I call the \emph{prevalence-inferred} approach takes a two-step approach.
First, it uses the statistical model from \cref{E-biology-data:sec:MRP} to estimate prevalence.
Then, it transforms the posterior distribution of prevalence to a posterior distribution of incidence using the forward substitution as in \cref{inc-prev:eq:forward-substitute}.
At a high-level, the statistical model is:
\begin{align}
    \vec{P}  &\dist F_P \\
    \rho &\dist F_\rho \\
    Y_{i,t} \mid \vec{P}, \rho &\dist \BB(n_{i,t}, P_{i,t}, \rho)
\end{align}
where $F_P$ is analogous to $F_Z$ in providing a joint prior for the prevalence in each stratum and day combination, and all other notation is as in the incidence-inferred approach.
The first step in the approach is to calculate the posterior distribution for $\vec{P}$.

The second step is to transform the posterior distribution for $\vec{P}$ to a posterior distribution for $\vec{Z}$.
The transformation is based on \cref{inc-prev:eq:forward-substitute}, which gives a one-to-one mapping between $\vec{P}$ and $\vec{Z}$; implementation details are in \cref{backcalc:sec:methods}.

\subsection{Desirable properties} \label{transmission:sec:backcalc-properties}

The incidence-inferred approach is more parsimonious in being a single step procedure.
However, computational considerations mean I favour the prevalence-inferred approach.

@@

The chosen approach should have the following properties.
\begin{enumerate}
    \item The prior should favour $\pi_{i,t}$ changing exponentially in $t$, \ie penalize changes from exponential growth.
        % Penalizing changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
        Such penalization has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately} and is the approach used for the CIS prevalence model (see \cref{E-biology-data:sec:MRP}).
    \item Limit estimates to epidemiologically and biologically plausible values.
        That is, incidence proportion and prevalence should be in the interval $[0, 1]$.
        % While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias.
    \item Borrow information across the different covariates.
        The borrowing should penalize changes in the distribution of incidence/prevalence across strata in the same geographic region.
        A constant distribution is predicted by the theoretical study of epidemics, for example the dynamics of stratified, mechanistic transmission models (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) when behavioural and immunity changes are negligible.
    \item Be computationally efficient.
        The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see \cref{E-biology-data:sec:SRS}).
\end{enumerate}

A parsimonious model satisfying properties 1 and 2 is to specify $\logit(\pi_{i,t}) = g_{\vec{\varphi}}(i, t)$, where $g$ is a function parameterised by ${\vec{\varphi}}$ and the prior on ${\vec{\varphi}}$ is specified to favour $d^2g/dt^2 = 0$, \ie favour linear growth in $\logit(\pi_{i,t})$.
The prior $F_Z$ or $F_P$ is now a prior on ${\vec{\varphi}}$.
The $\logit$ link function both limits $\pi_{i,t}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{i,t}$~\autocite{ealesAppropriately}.

The logit transformation also allows property 4 to be conveniently satisfied using an additive model for $g_{\vec{\varphi}}$.
An additive model on the logit scale is approximately multiplicative on the natural scale for small values of $\pi_{i,t}$ (which will be common).

@@

Properties 3 and 5 suggest the use of an approximate method for computing the posterior distribution.
A particular issue is that using MCMC, sufficient covariates to adjust for the non-response bias, and a period of more than 6--8 weeks leads to a model that is computationally prohibitive ~\citePersonalComms{Koen Pouwels}.
Instead, previous MRP implementations for the CIS used INLA (see \cref{E-biology-data:sec:MRP}).

INLA-based inference can only be applied to LGMs (\emph{Latent Gaussian Models}); see \cref{E-transmission:sec:INLA} for further details on INLA and LGMs.
The LGM framework requires that a transformation of $P_{i,t}$ is linear in $\vec{g} = [g_{{\vec{\varphi}}}(1, 1), \dots, g_{{\vec{\varphi}}}(1, T), \dots, g_{{\vec{\varphi}}}(2, 1), \dots, g_{{\vec{\varphi}}}(I, T)$ where $I$ and $T$ are the number of strata and days respectively.

If $\pi_{i,t} = P_{i,t}$ then this is trivially satisfied.
However, if $\pi_{i,t} = Z_{i,t}$, then no such linear relationship exists.
In particular:
\begin{align}
    P_{i,t}
    &= \sum_{k=0}^{\dmax} \pi_{i,t-k} S(k+1) \\
    &= \sum_{k=0}^{\dmax} \expit(g_{\vec{\varphi}}(i, t-k)) S(k+1).
\end{align}

@@

Now I explain how I extended the MRP model introduced in \cref{E-biology-data:sec:MRP} to implement the prevalence-inferred approach.
I start with the prevalence posterior from \cref{E-biology-data:sec:positivity-results}, denoting the times the estimates are for as $t = 1, \dots, T$.
I draw 500 samples of $\vec{P}$ from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).

Then, I transform this posterior sample using \cref{transmission:alg:calculate-Z}.
The algorithm is based on the forward-substitution algorithm from \cref{inc-prev:eq:forward-substitute}, which including stratum is:
\begin{align}
Z_{i,t}
&= P_{i,t} - \sum_{j=1}^{t-1} S(t - j + 1) Z_{i,j}.
\label{transmission:eq:forward-substitute}
\end{align}
Dividing each side of \cref{inc-prev:eq:forward-substitute} by the total number of individuals in each stratum transforms the number of prevalent individuals to prevalence and incidence to incidence proportion.
The algorithm uses a deterministic relationship between incidence and prevalence within a stratum, as \cref{E-inc-prev:sec:observation-process} justifies; \ie, the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.

@@

For each posterior sample of $\vec{Z}$, I poststratify it to the CIS target population and subpopulations of interest.
Specifically, define a (sub) population of interest as a set of strata $\set{S}$.
Then the incidence proportion in the subpopulation on day $t$ is
\begin{align}
Z_t(\set{S})
&= \frac{\sum_{j \in \set{S}} Z_{j,t} N_j}{\sum_{j \in \set{S}} N_j}
\label{backcalc:eq:poststratify}
\end{align}
where $N_j$ is the population size of stratum $j$.
The poststratified samples are now from the posterior of interest.

@@

A slow increase in incidence is estimated to occur over the autumn 2020 period, peaking in October in most regions (see \cref{transmission:fig:backcalc-regions}).
An exception is the North East and East of England regions (see \cref{transmission:fig:backcalc-regions}(C) and (D) respectively); these regions show no consistent increase or decrease in incidence during October.
All regions then decrease during November; partially coincidng with the national lockdown in force in England between 5 Nov and 2 Dec 2020~\autocite{ifgTimeline}.

During November, incidence starts rising, with regional heterogeneity in the timing of when this occurs (see \cref{transmission:fig:backcalc-regions}).
The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
There is weak evidence that when incidence starts increasing is correlated with the timing of the Alpha variant's spread (see \cref{transmission:fig:backcalc-alpha}).
A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
\begin{figure}
    \includegraphics{transmission/backcalc-regions}
    \caption[Incidence estimated using backcalculation by region]{%
        Incidence proportion for (A) all of England and (B) its component regions.
    }
    \label{transmission:fig:backcalc-regions}
\end{figure}
\begin{figure}
    \includegraphics{transmission/backcalc-alpha}
    \caption[Correlation of the emergence of Alpha and minimum incidence.]{%
        x-axis: time that S-gene target failure, a marker of the Alpha variant, is estimated to start increasing from baseline by region among CIS test results; from \textcite[table S3]{walkerTracking}.
        y-axis: posterior median and 95\% CrI of the date of minimum incidence in November or early December in each region.
        Linear regression line shown.
    }
    \label{transmission:fig:backcalc-alpha}
\end{figure}

@@

An edge effect, a rapid rise in incidence between days one and two, is visible (see \cref{transmission:fig:backcalc-start-effect}).
This is due to the incorrect assumption that incidence is constant in the past.
Changing this assumption (\eg to exponential growth) modifies the estimated value on the first day, but the rest of the series is insensitive to this change (not shown).
Therefore, the first day's estimate is unreliable, and has been excluded from all other results.
However, the rest of the series is robust to this assumption.
\begin{figure}
    \centering \includegraphics{transmission/backcalc-start-effect}
    \caption[Edge effects in backcalculation method]{%
        First seven days of incidence proportion estimate for the whole of England, highlighting the edge effect.
    }
    \label{transmission:fig:backcalc-start-effect}
\end{figure}

@@

The second set of assumptions are in specifying the prior for $\vec{P}$.
In particular, the prior places non-zero mass on prevalence values which imply negative incidence under the other model assumptions.
Rectifying this assumption is challenging without switching to the inference-inferred approach.
The inference-inferred approach could also allow more interpretable priors to be used because the infection process is the process that corresponds to biological and behavioural processes.
The inference-inferred approach, and other more powerful modelling approaches, require an alternative inference technique allowing a broader set of models than LGMs.

@@

I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcite{vanleeuwenTime,vanleeuwenAugmenting}.
%; these estimates have been used for previous modelling of the SARS-CoV-2 pandemic~\autocite{vanleeuwenTime,birrellRealtime}.
% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
This method starts with pre-pandemic baseline estimates of $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UKTUS (UK Time Use Survey)~\autocite{UKTUS}.
POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
UKTUS asked a representative sample of the UK population about how much time they spent doing various activities (\eg shopping and spending time at home) each day.
\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
Google mobility data are based on the location of Android users over time, relative to a pre-pandemic baseline.
The mobility and school data are averaged over a week down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
$\kappa_{aa'}(t)$ is assumed to be piecewise constant, changing only at the start of each week.
For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.

@@

The time-varying component, $\beta(t)$, allows for behavioural changes that impact transmission but are not captured by the contact matrices (\eg mask wearing).
I impose a random walk prior on $\log\beta(t)$ on the weekly timescale, smoothing its value over time.
This prior encodes that behaviour changes smoothly over time.
\begin{align}
    \beta(t) &= \beta_{\tilde{w}(t)} \\
    \log\beta_{w+1} &= \log\beta_w + \sigma_\epsilon \epsilon_w \\
    \epsilon_w &\sim \N(0, 1)
\end{align}
where $\tilde{w}(t)$ is the week number of time $t$, with week 1 being the first week considered.
Two hyperparameters will be inferred: $\beta_1$, the initial value of the random walk; and $\sigma_\epsilon$, the standard deviation of the random walk (see \cref{SEIR:sec:priors}).
$\beta(t)$ is the only time-varying fitted parameter.
Therefore, in practice, it will also absorb other effects such as any error in $\matr{\kappa}$.

@@

The remaining details of the transmission model structure follow \textcite{birrellRealtime}.
The transmission model is an SEIR model with two latent and two infectious states.
To form whole of England estimates, each region is modelled independently with the results aggregated using the poststratification methodology explained in \cref{backcalc:sec:methods}.

A diagram of the model for a single stratum is shown in figure \cref{SEIR:fig:full-model}.
The full system of differential equations for each region is below.
\begin{equation}
\begin{aligned}
    \vec{\lambda}(t) &= (\matr{\kappa}(t) \circ \matr{q}(t) \circ \matr{N}) ((\vec{i}_1(t) + \vec{i}_2(t))) \\
    \frac{d\vec{s}(t)}{dt} &= -\vec{\lambda}(t) \circ \vec{s}(t) \\
    \frac{d\vec{e}_1(t)}{dt} &= \vec{\lambda}(t) \circ \vec{s}(t) - 2\sigma \vec{e}_1(t) \\
    \frac{d\vec{e}_2(t)}{dt} &= 2\sigma \vec{e}_1(t) - 2\sigma \vec{e}_2(t) \\
    \frac{d\vec{i}_1(t)}{dt} &= 2\sigma \vec{e}_2(t) - 2\gamma \vec{i}_1(t) \\
    \frac{d\vec{i}_2(t)}{dt} &= 2\gamma \vec{i}_1(t) - 2\gamma \vec{i}_2(t) \\
    \frac{d\vec{r}(t)}{dt} &= 2\gamma \vec{i}_2(t) \\
    \frac{d\vec{p}_0(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \omega_0 \vec{p}_0(t) \\
    \frac{d\vec{p}_1(t)}{dt} &= \omega_0 \vec{p}_0(t) - \omega_1 \vec{p}_1(t) \\
    \frac{d\vec{p}_2(t)}{dt} &= \omega_1 \vec{p}_1(t) - \omega_{2} \vec{p}_2(t) \\
    \frac{d\vec{p}_3(t)}{dt} &= \chi \omega_{2} \vec{p}_2(t) - \omega_{3} \vec{p}_3(t)
\end{aligned}
\label{SEIR:eq:fullODEs}
\end{equation}
where $\circ$ indicates element-wise multiplication and $\matr{N}$ is a matrix with $(\matr{N})_{aa'} = N_{a'} / N_{a}$.
See \cref{SEIR:eq:fullODEs}

@@

Using the results of \cref{E-SEIR:sec:structured-populations}, $\R$ can also be written as $R_{\beta} = \beta_1 D(\matr{\kappa}(t) \circ \matr{q} \circ \matr{N})$ (recall that $D(\matr{\beta})$ denotes the dominant eigenvalue of $\matr{\beta}$).
Hence, I set $\beta_1 = R_\psi / D(\matr{\kappa}(t) \circ \matr{q} \circ \matr{N})$ to give $R_\psi = R_\beta$.
